A tutorial on Generalised Additive Mixed Effects Models for bilingualism research

### 2024-05-27 manuscript/manuscript.qmd
add summary and expand on fs

Readers interested in a more detailed conceptual discussion of GAMMs are referred to @soskuthy2017. The rest of this tutorial focuses on applying GAMMs in bilingualism research using the statistical programming language R [@rct2023]. While completing this tutorial will allow you to jump-start into fitting basic GAMMs, we invite interested readers to peruse the tutorials mentioned above to avoid unnecessary repetition of content here. Finally, we opted not to treat statistical significance testing in GAMMs (significance testing is treated in details in the aforementioned tutorials).

@@

Before looking at the summary of the `gam_1` model, we plot the predicted effect of proficiency. For this, we will use the tidygam package, which provides users with utility functions that make extracting and plotting predictions from GAMs more accessible.

@@

```{r}
#| label: gam-2-preds

@@

We could report this model like so: We fitted a Generalised Additive Model to learner accuracy scores, with a smooth term over proficiency (to model non-linear effects of proficiency) and a smooth term over time (to model non-linear effects of time). We also included a random intercept smooth for subject (to account for overall difference in mean accuracy by subject). According to the model, the effect of proficiency is non-linear and significant (*F* = 4638.38, *p* \< 0.0001). Based on the prediction plot, we observe that at lower proficiency, learning scores initially decrease and then start increasing again when proficiency reaches approximately 0.5. While the model summary indicates a significant effect of time on accuracy, the effect is negligibly small as it can be observed in the prediction plot of time. To summaries, the model suggests a non-linear effect of proficiency on learners' accuracy scores, but no effect of time.

@@

To include interactions between smooth terms, we can simply construct a new predictor with the combination of vowel and group and use that as the `by`-variable in the smooth terms. Due to how smooth terms are constructed when you include a `by`-variable, it is necessary to also include the variable as a parametric term (i.e. a classical linear term).

@@

>>>>

Readers interested in a more detailed conceptual discussion of GAMMs are referred to @soskuthy2017. The rest of this tutorial focuses on applying GAMMs in bilingualism research using the statistical programming language R [@rct2023]. While completing this tutorial will allow you to jump-start into fitting basic GAMMs with the mgcv package [@wood2006, @wood2011], we invite interested readers to peruse the tutorials mentioned above to avoid unnecessary repetition of content here. Finally, we opted not to treat statistical significance testing in GAMMs (significance testing is treated in details in the aforementioned tutorials).

@@

Before looking at the summary of the `gam_1` model, we plot the predicted effect of proficiency. For this, we will use the tidygam package [@coretta2023], which provides users with utility functions that make extracting and plotting predictions from GAMs more accessible.

@@

```{r}
#| label: gam-2-preds
#| warning: false

@@

We could report this model like so: We fitted a Generalised Additive Model to learner accuracy scores, with a smooth term over proficiency (to model non-linear effects of proficiency) and a smooth term over time (to model non-linear effects of time). We also included a random intercept smooth for subject (to account for overall difference in mean accuracy by subject). According to the model, the effect of proficiency is non-linear and significant (*F* = 4638.38, *p* \< 0.0001). Based on the prediction plot, we observe that at lower proficiency, learning scores initially decrease and then start increasing again when proficiency reaches approximately 0.5. While the model summary indicates a significant effect of time on accuracy, the effect is negligibly small as it can be observed in the prediction plot of time. To summarise, the model suggests a non-linear effect of proficiency on learners' accuracy scores, but no effect of time.

@@

To include interactions between smooth terms, we can simply construct a new predictor with the combination of vowel and group and use that as the `by`-variable in the smooth terms. Due to how smooth terms are constructed when you include a `by`-variable, it is necessary to also include the variable as a parametric term (i.e. a classical linear term). We would also like to allow for the modelled trajectories to vary by subject both in height and shape: while random effects included with `bs = "re"` (which we used in the first case study) account for difference in height and rotation of the fitted curves, *random factor smooth interaction* terms fit a different curve to each level of the chosen factor (like in our case, subject). Random factor smooth interaction terms are specified with `bs = "fs"`: the first argument in `s()` should be the variable to smooth over (here, `timep`) and the second argument is the factor (here `subj`) each of which levels will be fitted separately. Finally, the `m` argument corresponds to the order of the smoothing penalty: higher orders lead to more smoothing (the default is `2`, a second-order penalty). Setting `m = 1` corresponds to a higher constraint on the smooths of each level in the random factor, and it is thus a safeguard against over-fitting.

@@

## Summary

This introductory tutorial covered the basics of fitting non-linear effects using Generalised Additive Models (GAMs) and they mixed-effects counterpart Generalised Additive Mixed Models (GAMMs) with the mgcv and tidygam packages in R.

Two literature-based simulated case studies were employed to illustrate how to fit and plot GAMs. Case study 1 was about U-shaped learning and focused on modelling non-linear effects with smooths (`s()` terms in R syntax) and on obtaining model predictions for plotting. Case study 2 illustrated a time-series analysis of euclidean distance trajectories of three vowels (/a, i, u/) of three bilingual groups (simultaneous English/Spanish, low-proficiency late, high proficiency late). This case study introduced readers to `by`- variables, used to model effects of categorical factors on smooth, including interactions between factors, and random factor smooth interactions, which fit random effects.

This tutorial on its own will not be sufficient to develop an adequate GAM analysis but constitutes a low-barrier entry to readers who are interested in learning about GAMs. We encourage those who wish to run a full GAM analysis to peruse the tutorials mentioned in the introduction: @soskuthy2017, @soskuthy2021, @pedersen2019, @tamminga2016, @wieling2018. Important topics that were not covered here are: selection of type and number of basis functions in smooths, autocorrelation and autoregressive models, statistical inference and other distribution families.
